---
title: "Getting Started with ribobaser"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ribobaser)
```

## Overview

Using  paired ribosome profiling (RIBO) and RNA-seq (RNA) count matrices, we demonstrate how to:

- impute zeros prior to log-ratio transforms with `zero_imputation()`;
- filter low-abundance genes via `select_genes()`;
- remove problematic samples using `select_samples()` 
- calculate translation efficiency with `te()`;
- summarise TE profiles across studies/cell lines using `aggregate_samples()`.
- remove principal components using `pc_remove()`
- generate a network matrix using `tec()`

## Example data

Load the provided count matrices and accompanying metadata, then create a manageable
subset for illustration. The subset contains a mixture of samples with varying periodicity
scores so that the sample filters remove at least one entry.

```{r example-data}
set.seed(3)

data("ribo_raw_human_cap_995", package = "ribobaser")
data("rnaseq_raw_human_cap_995", package = "ribobaser")
data("Ribobase_QC_non_dedup_data", package = "ribobaser")
data("Ribobase_QC_dedup_data", package = "ribobaser")

qc <- Ribobase_QC_non_dedup_data
available_samples <- intersect(colnames(ribo_raw_human_cap_995), qc$Experiment)

# Choose four low-periodicity and four high-periodicity experiments to showcase filtering
ordered_idx <- order(qc$`Periodicity score`)
low_candidates <- qc$Experiment[ordered_idx][seq_len(4)]
high_candidates <- qc$Experiment[rev(ordered_idx)][seq_len(4)]
selected_samples <- intersect(c(low_candidates, high_candidates), available_samples)

# Restrict to a small gene set for faster examples
selected_genes <- sample(rownames(ribo_raw_human_cap_995), 400)

ribo_counts <- ribo_raw_human_cap_995[selected_genes, selected_samples, drop = FALSE]
rna_counts  <- rnaseq_raw_human_cap_995[selected_genes, selected_samples, drop = FALSE]

head(ribo_counts[, 1:3])
```

## Impute zeros with `zero_imputation()`

 The helper function `zero_imputation()` replaces zeros with ones across both matrices while retaining the original shape and non-zero entries. We have two options for zero imputation. 

```{r zero-imputation}
imputed <- zero_imputation(ribo_counts, rna_counts, method = "simple")
min(imputed$ribo)
```

```{r}
imputed_mult <- zero_imputation(ribo_counts, rna_counts, method = "multiplicative")
min(imputed_mult$ribo)
```

## Filter genes with `select_genes()`

We remove low-count genes by applying a CPM threshold. Setting
`dummy = TRUE` keeps the total count of removed genes in a synthetic
`DUMMY_REMOVED` row that downstream steps can inspect.

```{r select-genes}
gene_selection <- select_genes(
  imputed$ribo,
  imputed$rna,
  cpm = 10,
  fraction = 0.8,
  dummy = TRUE
)

tail(rownames(gene_selection$ribo), 3)
```

## Filter samples with `select_samples()`

Next we filter out samples that fail one of three criteria:

1. The dummy row accounts for more than 25% of remaining counts. GSM5598098 is removed by this filter
2. Periodicity score (from `Ribobase_QC_non_dedup_data`) falls below 0.55.
3. The CLR-transformed RNA counts explain less than 30% of the variance in clr(RIBO) when regressed  (`min_r2 = 0.3`). GSM5598098 also fails this filter.

```{r select-samples}
sample_selection <- select_samples(
  gene_selection$ribo,
  gene_selection$rna,
  high_dummy_percentage = 0.15,
  min_periodicity = 0.55,
  min_r2 = 0.3
)

colnames(sample_selection$ribo)
```

## Compute translation efficiency with `te()`

We enable two ways to calculate TE. We also have the option to run principal component removal on the Ribo and RNA matrices

```{r te-logratio}
te_logratio <- te(t(sample_selection$ribo), t(sample_selection$rna), method = "logratio", pc_removal = FALSE)

dim(te_logratio)
```

```{r te-regression, eval = requireNamespace("propr", quietly = TRUE) && requireNamespace("compositions", quietly = TRUE)}
te_regression <- te(t(sample_selection$ribo), t(sample_selection$rna), method = "regression", pc_removal = FALSE)
head(te_regression[, 1:3])
```

```{r te_pcremoved_pre}
te_pcremoved_pre <- te(t(sample_selection$ribo), t(sample_selection$rna), method = "regression", pc_removal = TRUE, n_pcs = 1)
head(te_pcremoved_pre[, 1:3])
```

## Aggregate TE profiles with `aggregate_samples()`

We average TE values per gene within each cell line. In this example, the three samples all come from the same cell line. 
The metadata input can either be "Ribobase_QC_dedup_data" or "Ribobase_QC_non_dedup_data", which include the metadata
for the Ribobase samples, or a metadata data.frame.

```{r aggregate-samples}
te_by_cell_line <- aggregate_samples(
  te_logratio,
  metadata = "Ribobase_QC_dedup_data",
  group_by = "cell_line"
)

head(te_by_cell_line)
```
## Principal component removal on the TE matrix with `pc_remove()`
In addition to the option of removing principal components on the Ribo and RNA matrices, principal components can also be removed from the TE matrix itself. PC removal can be run on a global scale or on a study-level scale. In this example, all 3 of the experiments are from the same study and therefore, pass the min_samples cutoff.

```{r pc-remove-global}
te_pcrem_global <- pc_remove(te_regression, method = 'global', n_pcs = 1)
head(te_pcrem_global$mat_pc[,1:3])
```
 
```{r pc-remove-study}
metadata <- "Ribobase_QC_non_dedup_data"
utils::data(list = metadata, package = "ribobaser", envir = environment())
metadata <- get(metadata, envir = environment())

te_pcrem_study <- pc_remove(te_regression, method = 'study', study_map = metadata, n_pcs = 1, min_samples = 3)
head(te_pcrem_study$mat_pc[,1:3])
```

## Calculating TEC based on TE with `tec()`

We can calculate gene networks based on TE. We have four methods of calculating networks: TEC (method = 'rho'), WGCNA (method = 'wgcna'), GLASSO (method = 'glasso'), and GENIE3 (method = 'genie3'). 

```{r tec-rho}
tec <- tec(te_regression, method = "rho")
head(tec[,1:5])
```

```{r tec-genie3}
tec <- tec(te_regression, method = "genie3")
head(tec[,1:5])
```

This sequence covers the primary features in *ribobaser*. In practice, you can adjust
thresholds and grouping variables to match experimental needs, and extend the pipeline
with custom visualizations or statistical modelling as required.
